---
title: "Storm Data Analysis"
author: "Den Lim"
date: "2022-11-24"
output: html_document
---


----------------------------------------------
** Synopsis



**Initial Set-up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
install.packages("tidyverse")
library(tidyverse)
```

**Loading csv using only relevant columns to preserve memory usage
```{r read csv, cache=TRUE}
  stormData <- read_csv("repdata_data_StormData.csv.bz2", 
                        show_col_types = FALSE, 
                        col_types=cols_only(EVTYPE=col_character(),
                                            FATALITIES=col_number(),
                                            INJURIES=col_number(),
                                            PROPDMG=col_number(),
                                            PROPDMGEXP=col_character(),
                                            CROPDMG=col_number(),
                                            CROPDMGEXP=col_character(),
                                            REFNUM=col_number()))
```

** Data processing
-use the PROPDMGEXP and CROPDMGEXP to create two new number columns
PROPDMGMUL and CROPDMGMUL to create multipliers for PROPDMG and CROPDMG
-replace multiplier NAs with 1's. 
-calculate totalDMG from the products of PROPDMGxPROPDMGMUL and CROPDMGxCROPDMGMUL
```{r processing}

newStormData <- stormData %>% 
                  mutate(PROPDMGMUL = ifelse(PROPDMGEXP=="M", 1000, 1),
                    CROPDMGMUL = ifelse(CROPDMGEXP=="M", 1000, 1)) %>%
                  select(EVTYPE, FATALITIES, INJURIES, PROPDMG, 
                         PROPDMGMUL, CROPDMG, CROPDMGMUL, REFNUM)
newStormData$PROPDMGMUL <- newStormData$PROPDMGMUL %>% 
                  replace_na(1)
newStormData$CROPDMGMUL <- newStormData$CROPDMGMUL %>% 
                  replace_na(1)
newStormData <- newStormData %>% 
                  mutate(totalDMG = (PROPDMG*PROPDMGMUL +
                                       CROPDMG*CROPDMGMUL))     
```
                     
